### 2025-06-06 chapters/background.tex

@@

Keystone is an open-source Trusted Execution Environment (TEE) framework designed specifically for the RISC-V architecture. It enables the development of secure enclaves that are customizable, lightweight, and suitable for both research and deployment in resource-constrained or security-critical environments. Keystone’s design reflects a modular approach, emphasizing minimalism in its Trusted Computing Base (TCB), portability across RISC-V platforms, and extensibility to accommodate advanced security features.

>>>>

\chapter{Background and System Architecture}
\label{chap:background}

The Keystone framework is built on the RISC-V instruction set architecture, which provides a unique foundation for secure system design due to its open specification and modular extensibility. Unlike traditional ISAs that are controlled by single vendors, RISC-V encourages community-driven enhancements and supports custom extensions, making it particularly attractive for trusted computing research and development. Its clean and well-documented privilege model allows for the implementation of isolation mechanisms that are critical to the construction of TEEs.

RISC-V defines three primary privilege levels: user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode). These privilege levels are hierarchical, with M-mode having the highest authority over system resources. M-mode has direct access to all hardware features and is responsible for critical functions such as interrupt control, exception handling, and access to physical memory protection mechanisms. Keystone leverages this privilege model by placing its Security Monitor (SM)—the core of the TCB—in M-mode. The SM is responsible for managing enclaves, enforcing memory isolation, and mediating access to sensitive system operations. By situating the TCB in M-mode, Keystone ensures that enclave isolation policies are enforced at the highest level of privilege, thereby minimizing the attack surface.

A central hardware feature used by Keystone is Physical Memory Protection (PMP), introduced in version 1.10 of the RISC-V Privilege Specification. PMP enables M-mode software to define access permissions for specific physical memory regions, controlling which lower-privilege modes (U and S) may read, write, or execute code in those regions. This capability forms the basis of Keystone’s memory isolation model. Each enclave is allocated a private memory region protected by PMP, ensuring that neither the host operating system nor any unauthorized software can access its contents. PMP rules are enforced directly by hardware, providing strong guarantees of isolation that are resistant to software-based attacks.

In addition to memory protection, RISC-V supports a flexible trap and exception handling mechanism. M-mode can intercept and handle all traps, but may delegate certain classes of exceptions—such as system calls and page faults—to S-mode for performance optimization. This delegation allows the enclave runtime to cooperate with the host operating system for memory management tasks, while still maintaining strict control over enclave boundaries. The virtual memory subsystem is managed through the standard RISC-V MMU, which translates virtual addresses to physical addresses using multi-level page tables. Keystone leverages this infrastructure to support enclaves that operate with their own private address spaces, protected by PMP from tampering or inspection.

Importantly, the use of standard programming models and toolchains for M-mode software development makes the Keystone Security Monitor both maintainable and extensible. Unlike microcode or hardwired logic, which are difficult to audit and update, the SM is implemented in C and can be subjected to conventional testing and formal verification techniques. This increases the transparency and trustworthiness of the TCB, which is a critical requirement for secure system design.

Taken together, the combination of RISC-V’s privilege hierarchy, PMP, trap delegation mechanisms, and extensibility provides a solid foundation for implementing robust and configurable TEEs. The Keystone framework capitalizes on these features to deliver a flexible and open platform for secure enclave execution, while remaining faithful to the RISC-V philosophy of openness and modularity.

@@

Keystone is an open-source Trusted Execution Environment (TEE) framework designed specifically for the RISC-V architecture. It enables the development of secure enclaves that are customizable, lightweight, and suitable for both research and deployment in resource-constrained or security-critical environments. Keystone’s design reflects a modular approach, emphasizing minimalism in its Trusted Computing Base (TCB), portability across RISC-V platforms, and extensibility to accommodate advanced security features. \cite{Lee2019}

### 2025-06-06 chapters/introduction.tex

>>>>

\chapter{Introduction}
Modern computing increasingly relies on platforms that operate in potentially untrusted environments. These include public cloud infrastructures, multi-tenant servers, and distributed edge devices, where sensitive data and critical computations may be exposed to compromised software stacks or malicious actors. In such contexts, ensuring the confidentiality and integrity of both data and execution is a fundamental requirement. Conventional security mechanisms, such as access control and encryption, are often insufficient in scenarios where the underlying operating system, hypervisor, or even firmware may be compromised.

To address these challenges, Trusted Execution Environments (TEEs) have emerged as a compelling solution. TEEs provide hardware-assisted isolation mechanisms that enable the execution of code in a protected context, referred to as an enclave. This enclave is designed to be resilient to attacks originating from untrusted system components, including the operating system and other user-space applications. By establishing a secure boundary around sensitive code and data, TEEs allow developers to implement trustworthy services even on compromised hosts.

Several commercial implementations of TEEs exist today, with Intel Software Guard Extensions (SGX) and ARM TrustZone being the most prominent examples. These technologies have demonstrated the practical viability of hardware-enforced isolation, and they have been integrated into a variety of real-world applications, from digital rights management to secure machine learning inference. However, despite their utility, these solutions are inherently proprietary and closed-source. This lack of transparency imposes significant constraints on researchers, developers, and system architects who wish to investigate alternative TEE designs, customize security mechanisms, or perform formal verification. Moreover, the architectural rigidity of these platforms limits their adaptability to emerging use cases, especially in academic or experimental contexts.

To overcome these limitations, the Keystone framework introduces an open-source, modular TEE architecture built atop the RISC-V instruction set architecture (ISA). RISC-V itself is an open, extensible ISA that has gained significant traction in both academic and industrial settings. Keystone leverages RISC-V's flexibility to enable fine-grained control over enclave behavior and to support a minimal, verifiable trusted computing base (TCB). The framework allows researchers and developers to explore custom enclave policies, experiment with hardware-software co-design, and perform reproducible evaluations in a controlled setting.

This thesis presents a detailed investigation into the Keystone TEE framework. It explores Keystone's architectural foundations, its use of RISC-V hardware features such as Physical Memory Protection (PMP), and its support for dynamic enclave management. Special attention is paid to the performance implications of enclave isolation, as measured through controlled benchmarking in a virtualized environment. By evaluating Keystone under varying execution conditions, this work seeks to provide a rigorous assessment of its efficiency and suitability for secure computation on emerging RISC-V platforms.

### 2025-06-06 main.tex

\documentclass[english, version-2020-11]{uzl-thesis}
\UzLStyle{computer modern oldschool design}
\usepackage{float}

\input{preamble.tex} % Put your \UzLStyle and \UzLThesisSetup here

@@

\chapter{Introduction}
\input{chapters/introduction.tex}

\chapter{Background and System Architecture}
The RISC-V instruction set architecture (ISA) serves as a promising foundation for TEE implementation due to its open and extensible nature. RISC-V offers unique benefits beyond its open-source status: it incorporates security-oriented primitives such as Physical Memory Protection (PMP) which facilitates efficient and fine-grained memory isolation. Moreover, its community-driven evolution fosters innovation in security features, enabling frameworks like Keystone to experiment with, and contribute to, ISA enhancements.

RISC-V defines three primary privilege modes: user (U-mode), supervisor (S-mode), and machine (M-mode). These modes govern the execution privileges and resource accessibility of software components. M-mode, the highest privilege level, has exclusive control over hardware resources and interrupt management, rendering it an ideal candidate for hosting the TCB, such as the Security Monitor (SM) in Keystone. Unlike hardware microcode, M-mode software can be developed and patched flexibly using standard programming languages and toolchains, improving maintainability and formal verification potential.

A key hardware feature leveraged by Keystone is the PMP, introduced in the RISC-V Privilege Specification version 1.10. PMP enables M-mode to enforce physical memory access restrictions on lower privilege levels, thus underpinning enclave memory isolation. Despite variations in PMP implementations across platforms, its guarantees are standardized, facilitating portable and secure enclave deployment.

Interrupt and exception handling in RISC-V is primarily managed at the M-mode level, which may delegate common traps, such as page faults and system calls, to S-mode to optimize performance. Virtual memory support is provided by a memory management unit (MMU) that translates virtual addresses to physical addresses using multi-level page tables. While the operating system controls virtual memory via the \texttt{satp} register, Keystone enclaves maintain protected page tables and are secured against OS-based attacks through PMP-enforced physical memory restrictions.

Collectively, these architectural features establish a robust foundation for implementing secure and efficient TEEs on RISC-V platforms, as exemplified by Keystone.

\input{chapters/background.tex}

\chapter{Methodology}
The objective of the performance study was to quantify the overhead introduced by Keystone's enclave isolation mechanisms on RISC-V architectures. Experiments were executed within a QEMU-based virtual environment configured for the RV64 architecture, providing consistency and hardware independence.

To run benchmarks, a separate enclave application (eapp) and corresponding host application were developed for each benchmark. The host application initialized and managed the enclave using the Keystone C++ API, while the enclave application contained the benchmark logic and communicated with the host through edge calls. This modular structure allowed each benchmark to be executed both natively (outside the enclave) and securely (within the enclave), enabling accurate measurement of the enclave-related overhead.
To achieve a comparative analysis, two execution modes were considered:
\begin{enumerate}
\item \textit{Native Execution}: Benchmark execution without enclave isolation.
\item \textit{Enclave Execution}: Benchmark execution within the Keystone TEE.
\end{enumerate}

Each benchmark—Dhrystone and CoreMark—was run multiple times under both conditions. Performance metrics, including total execution time, throughput (DMIPS or IPS), and result variability (standard deviation), were recorded and analyzed.

The evaluation focused on determining the relative performance degradation attributable to enclave isolation by comparing results across the two execution modes.

\input{chapters/methodology.tex}

\chapter{Results and discussion}
\input{chapters/results.tex}

\chapter{System Configuration Recommendations}
\input{chapters/recommendations.tex}

\chapter{Conclusion and Future Work}
\input{chapters/conclusion.tex}

\begin{bibtex-entries}
@TechReport{Kernighan1974,
author = {Brian Kernighan},
title = {Programming in C – A Tutorial},
institution = {Bell Laboratories},
year = {1974}
}
\end{bibtex-entries}
\end{document}

>>>>

\documentclass[english, version-2020-11]{uzl-thesis}
\UzLStyle{pagella contrast design}
\usepackage{comment}
\usepackage{float}

\input{preamble.tex} % Put your \UzLStyle and \UzLThesisSetup here
\addbibresource{refs.bib}

@@

%\chapter{Introduction}

\include{chapters/introduction.tex}

\include{chapters/background.tex}

\include{chapters/methodology.tex}

\include{chapters/results.tex}

\include{chapters/recommendations.tex}

\include{chapters/conclusion.tex}

\end{document}
